Newton's Method in Machine Learning

### 2025-04-09 latex/Draft/Draft.tex
Changes
changes

\begin{document}
\maketitle
\section{Introduction}
In this project, we explored and tested the practical effectiveness of Newton's Method as an optimization technique within the machine learning context. We were interested in investigating whether this second-order method could be effectively adapted and optimized to train neural networks on a simple supervised learning problem: digit classification using the digits dataset.

@@

We must prove Taylor's theorem for the high dimensional case. We shall do so directly.
Let $f : \mathbb{R}^n \mapsto \mathbb{R}$. We shall observe some properties of $f$ in the region of $A$, where $A$ is an open subset of $\mathbb{R}^n$. We shall make the following assumptions: $f$ is continuous on $A$, $f$ is differentiable on $A$. (Note: We do not contain the boundary points as this allows us to handle asymptotics more neatly). 
Let us define the derivative of $f$ in the direction of $\alpha$ by writing \begin{equation}\vec{\alpha} \cdot \vec{\nabla} f(\vec{x}) = \lim_{t \to 0}\frac{f(\vec{x} + \vec{\alpha}t) - f(\vec{x})}{t}\end{equation}, under the assumption that $\vec{x} + t \vec{\alpha} \in A$. This allows us to find the derivative in any direction from the location of $x$. Let us define $\Phi(t) = f(\vec{x} + \vec{\alpha} t)$. Because $\Phi : \mathbb{R} \mapsto \mathbb{R}$
we can use the one dimensional form of Taylor's theorem \cite{wikipediacontributors-2019-taylor}on $\Phi$. This lets us write the equivalency 
\begin{equation}\Phi(t) = \Phi(0) + \Phi'(0)t + \Phi''(0)\frac{t^2}{2} + \dots\end{equation}
By substituting the definition of $\Phi(t)$ into both sides, we can write 
\begin{equation} f(\vec{x} + t \cdot \vec{\alpha}) = \Phi(0) + \Phi'(0)t + \Phi''(0)\frac{t^2}{2} + \dots\end{equation}
We also know that 
\begin{equation}\Phi(t)' = \lim_{t \to 0}\frac{f(\vec{x} + \vec{\alpha}t) - f(\vec{x})}{t} = \vec{\alpha} \cdot \vec{\nabla}f(\vec{x})\end{equation}
By the definition of $\Phi$, and $\vec{\alpha} \cdot \vec{\nabla}$.
By substituting into both sides, we get the following. For convenience, we shorten $\vec{\alpha}\cdot\vec{\nabla}$ to $\vec{\nabla}$.
\begin{equation} f(\vec{x} + t \cdot \vec{\alpha}) = f(\vec{x}) + [\vec{\nabla}f(\vec{x})]^T\vec{\alpha} t + \frac{1}{2}\vec{\alpha}^T[\vec{\nabla}^2f(\vec{x})]\vec{\alpha} \cdot t^2 + \dots\end{equation}
We shall only use first and second order methods, so we have not shown the third order term of the Taylor Series, which requires rank 3 tensor multiplications.
Thus we have proved Taylor's Theorem for functions $f : \mathbb{R}^n \mapsto \mathbb{R}$, under an open subset of $\mathbb{R}^n$.

@@

We shall now study the convergence of algorithms of the following form $x_{k+1} = x_k - S_k\vec{\nabla}f(x_k)$. The constraints we place are that $S_k\vec{\nabla}f(x_k)$ is sufficiently small as to not overshoot the extrema, and that $S_k$ is a positive definite matrix. For example, Newton's method chooses $S_k = \vec{\nabla}^{-2}f(x_k)$.

@@

\section{Network model}
We used Tensorflow with Keras to build our network model. The dataset we used was "Pen-Based Recognition of Handwritten Digits" which consists of 8x8 images of handwritten digits. Our network consists of an input layer of 64 neurons (one for each pixel), a hidden layer of 32 neurons with ReLU as their activation function, and an output layer of 10 neurons (one for each digit) with softmax as the activation. For the ADAM and SGD optimizers, we used Keras' built-in optimizers. We wrote the Newton's Method optimizer ourselves using Keras' automatic differentiation (through the use of Gradient Tapes and the built-in Jacobian method). The Adam and SGD models were trained using Tensorflow's fit method, and the Newton's method model was trained using our own train function, as there was no pre-existing function to do this. When analyzing the results, it is important to take into account that the networks using ADAM and SGD will have a slight advantage over the network using our custom Newton's Method optimizer, as the code for these two optimizers was written and optimized by Tensorflow developers over many years.

>>>>

\begin{document}
\maketitle
\section{Introduction}
In this project, we explored and tested the practical effectiveness of Newton's Method as an optimization technique within the machine learning context. We were interested in investigating whether this second-order method could be effectively adapted and optimized to train neural networks on a simple supervised learning problem: digit classification using the digits dataset \cite{digits}.

@@

We must prove Taylor's theorem for the high dimensional case. We shall do so directly.
Let $f : \mathbb{R}^n \mapsto \mathbb{R}$. We shall observe some properties of $f$ in the region of $A$, where $A$ is an open subset of $\mathbb{R}^n$. We shall make the following assumptions: $f$ is continuous on $A$, $f$ is differentiable on $A$. (Note: We do not contain the boundary points as this allows us to handle asymptotics more neatly). 
Let us define the derivative of $f$ in the direction of $\alpha$ by writing \begin{equation}\vec{\alpha} \cdot \vec{\nabla} f(\vec{x}) = \lim_{t \to 0}\frac{f(\vec{x} + \vec{\alpha}t) - f(\vec{x})}{t}\end{equation}, under the assumption that $\vec{x} + t \vec{\alpha} \in A$. This allows us to find the derivative in any direction from the location of $x$. Let us define $\Phi(t) = f(\vec{x} + \vec{\alpha} t)$. Because $\Phi : \mathbb{R} \mapsto \mathbb{R}$
we can use the one dimensional form of Taylor's theorem \cite{wikipediacontributors-2019-taylor}on $\Phi$. This lets us write the equivalency 
\begin{equation}\Phi(t) = \Phi(0) + \Phi'(0)t + \Phi''(0)\frac{t^2}{2} + \dots\end{equation}
By substituting the definition of $\Phi(t)$ into both sides, we can write 
\begin{equation} f(\vec{x} + t \cdot \vec{\alpha}) = \Phi(0) + \Phi'(0)t + \Phi''(0)\frac{t^2}{2} + \dots\end{equation}
We also know that 
\begin{equation}\Phi'(t) = \lim_{t \to 0}\frac{f(\vec{x} + \vec{\alpha}t) - f(\vec{x})}{t} = \vec{\alpha} \cdot \vec{\nabla}f(\vec{x})\end{equation}
By the definition of $\Phi$, and $\vec{\alpha} \cdot \vec{\nabla}$.
By substituting into both sides, we get the following. For convenience, we shorten $\vec{\alpha}\cdot\vec{\nabla}$ to $\vec{\nabla}$.
\begin{equation} f(\vec{x} + t \cdot \vec{\alpha}) = f(\vec{x}) + [\vec{\nabla}f(\vec{x})]^T\vec{\alpha} t + \frac{1}{2}\vec{\alpha}^T[\vec{\nabla}^2f(\vec{x})]\vec{\alpha} \cdot t^2 + \dots\end{equation}
Note: The Hessian $\vec{\nabla}^2f(\vec{x})$ appears here because it is the multidimensional equivalent of $\Phi''$. In other words, the Hessian appears here because it encodes the second order information required by the Taylor Polynomial.
We shall only use first and second order methods, so we have not shown the third order term of the Taylor Series, which requires rank 3 tensor multiplications.
Thus we have proved Taylor's Theorem for functions $f : \mathbb{R}^n \mapsto \mathbb{R}$, under an open subset of $\mathbb{R}^n$.

@@

We shall now study the convergence of algorithms of the following form $x_{k+1} = x_k - S_k\vec{\nabla}f(x_k)$. The constraints we place are that $S_k\vec{\nabla}f(x_k)$ is sufficiently small as to not overshoot the extrema, and that $S_k$ is a positive definite matrix. For example, Newton's method chooses $S_k = \vec{\nabla}^{-2}f(x_k)$ \cite{recht-wright}.

@@

\section{Network model}
We used Tensorflow with Keras to build our network model. The dataset we used was "Pen-Based Recognition of Handwritten Digits"\cite{digits} which consists of 8x8 images of handwritten digits. Our network consists of an input layer of 64 neurons (one for each pixel), a hidden layer of 32 neurons with ReLU as their activation function, and an output layer of 10 neurons (one for each digit) with softmax as the activation. For the ADAM and SGD optimizers, we used Keras' built-in optimizers. We wrote the Newton's Method optimizer ourselves using Keras' automatic differentiation (through the use of Gradient Tapes and the built-in Jacobian method). The Adam and SGD models were trained using Tensorflow's fit method, and the Newton's method model was trained using our own train function, as there was no pre-existing function to do this. When analyzing the results, it is important to take into account that the networks using ADAM and SGD will have a slight advantage over the network using our custom Newton's Method optimizer, as the code for these two optimizers was written and optimized by Tensorflow developers over many years.
